---
title: "Estimating Survival Rate of Titanic Passengers"
author: "Lemuel Kumarga"
always_allow_html: yes
knit: (function(inputFile, encoding) { source("shared/knit.R"); knitter(inputFile, encoding)})
---

```{r echo=FALSE, warning=FALSE, results='hide'}
packages <- c("knitr")
tmp <- lapply(packages, library, character.only = TRUE)

read_chunk("main.R")
read_chunk("shared/helper.R")
  
```

## Problem Description

Set for its maiden voyage in 1912, RMS Titanic was then considered unsinkable due to its robust structure. After picking up passengers in Southampton, Cherbourg and Queenstown, the ship set sail towards New York. It was on their way there that the Titanic met its calamitous fate, crashing into an Iceberg and rendering more than 50% of the passengers lost.

There were many questions concerning the events leading to and during the disaster. Some of these include: Which part of the ship was most damaged? What was the evacuation process of the passengers? Using data containing the personal details and disaster outcomes (survival/lost) of passengers, we aim to <b>discover the various factors affecting the survival likelihood of passengers.</b> 

<span class="color-1-text">FYI: This problem was posed as an introductory challenge in <a href="https://www.kaggle.com/c/titanic" target="_blank">Kaggle</a>. The passengers are divided into two sets, one used for training and the other for testing. The training sets contain both the personal details of passengers and their outcomes, while the testing set only contain the former. By developing a robust model through the training set, we hope to predict the survival likelihood of passengers in the testing set.</span>

## Summary of Results

To be completed.

## Preliminaries

First load the necessary packages for this exercise.
```{r}
<<init>>

si <- sessionInfo()
base_pkg_str <- paste0("Base Packages: ",paste(si[["basePkgs"]], collapse=", "))
attached_pkg_str <- paste0("Attached Packages: ",paste(names(si[["otherPkgs"]]), collapse=", "))
cat(paste0(base_pkg_str,"\n",attached_pkg_str))
```

## Exploration

### About the Data

We kick off by exploring the data that was provided:
```{r}
<<data_overview>>
  
<<data_comp>>

pander(cols_summary, caption='Titanic Passengers Data - For more info, please visit <a href="https://www.kaggle.com/c/titanic/data" target="_blank">Kaggle</a>')
```
<br>

Based on the above table, notice the following:

* <span class="hl color-1-text">Names</span> are aggregated in the following format: <span class="hl color-2-text">Last_Name, Title 
First_Name</span>.
* <span class="hl color-1-text">Passenger class (Pclass) </span> levels are the equivalent of <span class="hl color-2-text">First Class (1), Business Class (2) and Economy Class (3)</span>. Hence, they could be used as an indicator of income level.
* There are different types of <span class="hl color-1-text">Ticket</span> formats.
* <span class="hl color-1-text">Age</span> has missing data to be populated.
* <span class="hl color-1-text">Cabin</span> data is sparse, and hence we may want to exclude this for the prediction model.

### Preliminary Insights

<div class="st">Insight 1: Wealthy individuals are more likely to survive.</div>

Using passenger class as a proxy for wealth, we calculated the survival likelihood (% of passengers survived) for each class.
```{r exp_income, fig.height=3, fig.width=10, fig.align="center"}

```
The chart above shows that the more premium the class, the more likely the passengers were to survive. One potential reason explaining this insight could be that 1st class passengers were the first in line to access the lifeboats, while 3rd class passengers were left to fend for themselves.

```{r exp_fare, fig.height=3, fig.width=10, fig.align="center"}

```
Similarly, we also noticed this phenomenon in fares, where the higher amount an individual paid for the fares, the more likely he/she will survive the crash.

<div class="st">Insight 2: Females and more esteemed individuals are more likely to survive than males.</div>

The data provides us with a list of titles under the <span class="color-1-text">Name</span> column. Each title is associated with a gender (with the exception of one), as shown below:

```{r exp_titles, fig.height=2.6, fig.width=10, fig.align="center"}

```

Other than <span class="hl color-1-text">Mr, Miss, Mrs and Master</span>, all other titles are not presumed by many passengers. To navigate the small sample size, we group all of these other titles under <span class="hl color-2-text">Rare Title</span>.

Using the following modifications, we can then assess the survival likelihood of each title and gender:

```{r exp_title_gender, fig.height=3, fig.width=10, fig.align="center"}

```

<span class="hl">All</span> females, regardless of their titles, have higher survival likelihoods than males. In addition, having a title other than Mrs, Miss or Mr, elevates the survival likelihood of the individual. One reason explaining this insight could be that females and esteemed individuals were given early access to escape the Titanic. 

<div class="st">Insight 3: The younger the passenger, the more likely he/she will survive.</div>

The chart below shows how survival likelihood changes according to age.

```{r exp_age, fig.height=3, fig.width=10, fig.align="center"}

```

With the limited data set, we can see that younger individuals are more likely to survive. A possible explanation is probably because babies and toddlers occupy less space (and hence easier to find seats within the lifeboats). 

However, older people, especially those more than 60 years old, are less likely to survive. This is probably due to their lack of agility in responding to the crash. 

<div class="st">Insight 4: An individual is more likely to survive if he/she has other family members with him/her.</div>

Typically, we would expect that the more family members there is in Titanic, the less likely an individual will survive. This is because individuals would likely go into lifeboats with their family members, making it harder to find space in the lifeboats.

However, contrary to expectations, the data shows that the larger the family size (individual + siblings/spouses + parent/children), the more likely an individual is likely to survive:

```{r exp_company, fig.height=4, fig.width=10, fig.align="center"}

```

As can be seen from the chart above, individuals with 1 to 2 children are more likely to survive than those without any children. It is inconclusive to determine from the chart above, whether same-age company has an impact of survivalhood. Hence, this premise is <span class="hl yellow-text">TRUE TO A LIMITED EXTENT</span>.

#### On Cabin Positions

<div class="st">Hypothesis 6: Cabin positions should have an impact of survivalhood, but only to a certain extent.</div>

Cabins further away from the lifeboats will have a more difficult time surviving, hence we should assume that where passengers stay will have an impact on their survivalhood. However, the cabin positions can only determine survivalhood partially, given that passengers may not be at their cabins during the time of crash.

A more detailed deck plan can be found <a href="https://www.encyclopedia-titanica.org/titanic-deckplans/b-deck.html" target="_blank">here</a>. As can be seen, the letter of the cabin represents the deck (floor) where the room is located. We also see that odd numbers correspond to one side of the Titanic, while even numbers correspond to the other.

We will break down the cabin position into three constituents: the cabin floor, the cabin number (odd or even), and the number of cabins specified.

```{r exp_hypo7}

```

##### Cabin Floors

```{r exp_hypo7_p1, fig.height=5, fig.width=7, warning=FALSE, fig.align="center"}

```

From the chart, we can deduce that <span class="hl yellow-text">Cabins B to E</span> has a higher likelihood of survival compared to other/unspecified cabins, which suggests that the position of cabins play a role in determining whether a passenger survive.

##### Cabin Numbers

```{r exp_hypo7_p2, fig.height=3, fig.width=10, fig.align="center"}

```

It is pretty clear that those who stay in the odd rooms are more likely to survive than those in the even rooms.

##### Cabins Specified

```{r exp_hypo7_p3, fig.height=5, fig.width=7, fig.align="center"}

```

Even though seems to be an inverse relationship between the survival likelihood and number of cabins specified, the sample size is too small to ensure statistical significance.

In conclusion, cabin floors and cabin numbers can determine a passenger's survival likelihood, while there is too little data to deduce a relationship in the number of cabins specified. Hence, the hypothesis is <span class="hl green-text">TRUE</span>.

#### On Embakartion Ports

<div class="st">Hypothesis 7: Port of embarkation should have no impact on survivalhood.</div>

```{r exp_hypo8, fig.height = 3, fig.align="center"}

```

<br>
At first glance, there seems to be no plausible explanation why a passenger's port of embarkation will have an impact on the passenger's survival likelihood.

```{r exp_hypo8_p2, fig.height = 3, fig.width=10, fig.align="center"}

```

However, by studying the demographics of the passengers who embarked at each port, we know that a higher proportion of Cherbough are high-income. This explains the higher survival likelihood for those who embarked at Cherbough, and renders the hypothesis to be <span class="hl red-text">FALSE</span>.

## Preparation

Based on the above hypotheses, we will be adding the following features to predict the survival likelihood of an individual.

Feature | Variable Type | Hypotheses Supporting the Feature  | Null Handling 
- | - | - | -
Pclass | Continuous | Hypothesis 1 | -1
Fare | Continuous | Hypothesis 1 | -1
Sex | Categorical | Hypothesis 3 | 'Unknown' Category Variable
Title | Categorical | Hypothesis 3 | 'Unknown' Category Variable
Age | Continuous | Hypothesis 4 | Regression
SibSp | Continuous | Hypothesis 5 | -1
Parch | Continuous | Hypothesis 5 | -1
CabinFloor | Categorical | Hypothesis 6 | 'Unknown' Category Variable
CabinNumber | Categorical | Hypothesis 6 | 'Unknown' Category Variable
Port of Embarkation | Categorical | Hypothesis 7 | 'Unknown' Category Variable

Below is an example of the cleansed data set for model ingestion.
```{r}
<<data_prep>>

pander(head(cleaned_set))
```

## The <i>Random Forest</i> Model

Listed below are the preliminary results from the initial run using the Random Forest model.

```{r data_model}

```


```{r}
<<data_model_err_anal>>
  
pander(head(err_out))

```

## Predictive Results
